\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-17 docs/paper/paper final.tex
Removed some comment statements.

%EMS-PC extension (no pseudo-code, but additions regarding standard PC)
\paragraph{EMS-PC}
We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
These three changes have been made in this adaption with respect to standard PC as described in algorithm~\ref{alg:pc_alg}.
At line~\ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
All of these separating sets are recorded.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}\label{sec:pipeline}
%- What type of data is needed
PC algorithm requires observational data where each instance is a vector of single observations per variable.
We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.

@@

% Comparison standard PC and EMS-PC
We compare standard PC and EMS-PC on their causal graph.
As EMS-PC is more constrained in concluding directionality, it may find less erroneous edges than standard PC.
As we cannot verify this directly, we compare the total count, consistency and standard-deviation of one-directional arrows.
%- Run-time of methods and comparison between run-times
We also compare run-time per iteration between both.

@@

\subsection{Structural graphs}
%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
%Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
%This is done both per subject, and as an avarage over all subjects.
%We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
Modified PC and EMS-PC do not differ in finding the structure.
As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
The results are presented per subject in figure~\ref{fig:struct_subjects} and as an avarage across all subjects in figure~\ref{fig:struct_avg}.
For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop} % TODO remove colorbar, why is it even there
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
  \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
\label{fig:struct_subjects}
\end{figure}

@@

\subsection{Directed graphs}
This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
As there were no notable differences between different subjects, only the avarage PDAG across all subjects is presented.
The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
Although many arrowheads are found with probabilities above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
  \label{fig:pdag_avg}
\end{figure}

Another representation of this data has been used to emphasise the asymmetric parts of the directed graphs.
To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}. The avarage across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
  \caption{The assymetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right. A high value denotes a high certainty of an one-directional arrow. The avarage of all six subjects is taken.}
  \label{fig:asym_avg}
\end{figure}

@@

\section{Discussion}
%Good consistency of structure within and between subjects
Many structural connections found by the PC algorithm are found consistently, both within one subject and also across multiple subjects.
The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus collosum.
The consistency can also be seen in the distribution of values in figure~\ref{fig:distributions}.
Non-zero values make up approximately 16 percentage points of the graph, of which 3 percentage points show a certainty larger than 0.5.
Approximately 1 percentage point even shows a certainty of 1.

>>>>

%EMS-PC extension
\paragraph{EMS-PC}
We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
These three changes have been made in this adaption with respect to standard PC as described in algorithm~\ref{alg:pc_alg}.
At line~\ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
All of these separating sets are recorded.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}\label{sec:pipeline}
% What type of data is needed
PC algorithm requires observational data where each instance is a vector of single observations per variable.
We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.

@@

% Comparison standard PC and EMS-PC
We compare standard PC and EMS-PC on their causal graph.
As EMS-PC is more constrained in concluding directionality, it may find less erroneous edges than standard PC.
As we cannot verify this directly, we compare the total count, consistency and standard-deviation of one-directional arrows.
% Run-time of methods and comparison between run-times
We also compare run-time per iteration between both.

\section{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.

All measurements and pre-processing as described in this section has been done during previous research. No additional pre-processing was needed in our research in order to apply the PC algorithm on this data.

@@

\subsection{Structural graphs}
% Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
% Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
% This is done both per subject, and as an avarage over all subjects.
% We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
Modified PC and EMS-PC do not differ in finding the structure.
As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
The results are presented per subject in figure~\ref{fig:struct_subjects} and as an average across all subjects in figure~\ref{fig:struct_avg}.
For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
  \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
\label{fig:struct_subjects}
\end{figure}

@@

\subsection{Directed graphs}
This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
As there were no notable differences between different subjects, only the average PDAG across all subjects is presented.
The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
Although many arrowheads are found with probabilities above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The average of all six subjects is taken.}
  \label{fig:pdag_avg}
\end{figure}

Another representation of this data has been used to emphasise the asymmetric parts of the directed graphs.
To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}. The average across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
  \caption{The asymmetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right. A high value denotes a high certainty of an one-directional arrow. The average of all six subjects is taken.}
  \label{fig:asym_avg}
\end{figure}

@@

\section{Discussion}
%Good consistency of structure within and between subjects
Many structural connections found by the PC algorithm are found consistently, both within one subject and also across multiple subjects.
The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus callosum.
The consistency can also be seen in the distribution of values in figure~\ref{fig:distributions}.
Non-zero values make up approximately 16 percentage points of the graph, of which 3 percentage points show a certainty larger than 0.5.
Approximately 1 percentage point even shows a certainty of 1.
